The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-05 kalman_decomposition.tex
edited through

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

\usepackage{multicol}

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

\newcommand{\calA}{\mathcal{A}}
\newcommand{\calH}{\mathcal{H}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

\DeclareMathOperator{\spn}{span}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\begin{document}

\section*{Evolutionary conservation of phenotype does not entail conservation of the underlying molecular mechanism leading to the rapid formation of new species}
{\centering
Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
$^{\dagger}$University of Southern California, Los Angeles, California \qquad $^{\ddagger}$University of Oregon, Eugene, Oregon \\
\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}
\\
}

%\begin{multicols}{2}

\begin{abstract}
We introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks. The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes. Here we give an exact expression for the set of all linear mechanisms with identical phenotypes, and expect evolution under neutrality to explore this set. We employ a quantitative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit. We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
\end{abstract}

  \plr{Additional ideas to consider adding:
\begin{itemize}
    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
    \item hybrid vigor (need to do calculation)
    \item discuss linearity, linearization, and canalization in introduction
    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
\end{itemize}}

%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}. This is especially pertinent given the staggering complexity and scope of contemporary research programs. This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice, as models allow the development of concrete numerical predictions. Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.

%Presently, we focus on the neutral evolution of genetic regulatory networks. That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space. This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.

%It is commonly taught that an organism's genome contains the heritable material 
%that natural selection filters and that an organism's phenotype directly determines its
%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
%undrstood molecular machinery -- and it is a major goal common to many different disciplines
%within the life sciences to elucidate its form, function, and evolution. These aims are
%delicately intertwined and a comphrehensive understanding of a system's evolution requires
%an understanding of its function and vice versa.

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary, draghi2015robustness} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
parameter schemes, or even different mechanics can be equally accurate, but still not
\emph{actually} agree with what is being modelled. In control theory, where electrical 
circuits and mechanical systems are often the focus, it is understood that there can be an 
infinite number of ``realizations'', or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Although this may frustrate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

  In this paper we outline a theoretical framework to study the evolution of biological systems. We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), however presently, we focus solely on neutral evolution, that is where phenotype is conserved over evolutionary time.

  We derive an analytical description of the set of all linear biological systems with identical phenotypes -- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift. In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

\section*{Gene Networks as Linear Dynamical Systems}

%  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. 
%   Thus we say that some function $\phi(t)$ is a phenotype where, 
%  \begin{align}
%    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
%  \end{align}
%  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study. The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity. 
%
%  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$. 
%
%  We describe the impulse response as, 
%  \begin{align}
%    h(t) = C e^{A t} B
%  \end{align}
%  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection. 
%
%  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$. 
%
%  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$, is defined as,
%  \begin{align}
%    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
%  \end{align}
%  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
%}
Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems. We define an organism's phenotype $\phi(t)$ to be the molecular time-dynamics directly relevant to survival. These dynamics are the result of the interconnections of a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) and an environmental input $u(t)$.
Such a system $\mathcal{S}$, is composed of two equations,
  \begin{align}
    \mathcal{S} := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
  \end{align}
 Where the \emph{kryptotype} $\kappa(t)$, is a list of a system's molecular concentrations at time $t$ -- or more generally the systems \emph{internal dynamics}. As some system dynamics may not be visible nor relevant to selection, we distinguish the \emph{kryptotype} (as it is hidden) from the \emph{phenotype} -- which is the systems \emph{external dynamics}.
  The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the systems current state (or kryptotype), the system architecture $A$ -- such as a gene regulatory network, the environmental input $u(t)$, and $B$ (an $n \times l$ matrix) -- how the system processes its environment. $C$ (an $l \times n$ matrix) filters the dynamics relevant to survival.
  Thus the phenotype is simply a convolution of the system organization and the environment, 
  \begin{align}
    \phi(t) = \int_{0}^{t} C e^{A t} B u(t) dt ,
  \end{align}
  where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system. As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regulatory element (or promoter) for gene $i$.

%We will now lay out the model in more general terms.
%Suppose that the \emph{internal state} of the system
%is parameterized by the concentrations of a collection of $n$ molecular species,
%$S_1, \ldots, S_n$,
%and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
%There are also $m$ ``input'' species, whose concentrations are determined
%exogenously to the system,
%and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
%and $\ell$ ``output'' species, whose concentrations are denoted
%$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
%The output is merely a linear function of the internal state:
%\begin{align*}
%    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
%\end{align*}
%Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
%and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
%as it is ``hidden'' from direct selection.
%The rate at which the $i^\text{th}$ species is produced
%is a weighted sum of the concentrations of the other species
%as well as the input:
%\begin{align*}
%    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
%\end{align*}
%In matrix notation, this is written more concisely as
%\begin{align} \label{eqn:lti_system}
%    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
%    \phi(t) &= C \kappa(t) .
%\end{align}

  \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}

    Cellular division is a complicated phenomena, governed by many different processes, however it is agreed that its rhythm is partially controlled by periodic (oscillating) gene transcription \citep{orlando2008global}. Consider a simplified model of oscillating gene transcription. In the present framework periodic expression requires at minimum two interacting genes.

    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates the transcription of gene-2 with equal magnitude (of $1$) and relative to each of their concentrations, denoted by $\kappa_{1}$ and $\kappa_{2}$. Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network). Lastly suppose that the production of both genes is stimulated by an impulse of a molecule present immediately after division.

    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
    \end{align*}
    where $\dot{\kappa}$ denotes the time derivative. The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time. If we record the regulatory coefficients in the matrix
    \begin{align*}
      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
    \end{align*}
    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
      \begin{align*}
      \dot{\kappa} &= A \kappa(t) + B u(t) .
      \end{align*}
      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
      \begin{align*}
      \phi(t) &= C \kappa (t)
      \end{align*}
      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. (Note: if the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}

      Since the input is simply an impulse, its phenotype is equivalent to its impulse response

      \begin{align*}
        \phi(t) = h(t) = \sin(t) + \cos(t)  .
      \end{align*}

    \begin{figure}[H]
       % \begin{center} 
      \centering
         \begin{tabular}{cc}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$\kappa_{1}$};
                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
                \node[shape=rectangle] (U) at (1.5,2) {input};
                \node[shape=rectangle] (y) at (1.5,-2) {output};
            \end{scope}

            \begin{scope}[>={Stealth[black]},
                          every node/.style={fill=white,circle},
                          every edge/.style={draw=black, thick}]
                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
            \end{scope}
            \begin{scope}[>={Stealth[black]},
                          every edge/.style={draw=black, thick}]
                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
            \end{scope}

            \end{tikzpicture} &
       % \end{center}
      %  \caption{
      %      Diagram of Example \ref{ex:osc} in the text.
      %      \plr{explain what arrows mean if nec}
      %      \label{fig:oscillator_diagram}}
   % \end{figure}
   % \begin{figure}[H]
    %  \centering
      \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_impulse}
   \end{tabular}
  %  \end{center}
      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.}
    \end{figure}
      We return to the evolution of such a system below.
  \end{example}

  \section*{Linear Evolutionary Systems}

  The literature is filled with detailed observations of molecular systems and their diversity. There are examples of significant diversity in the networks underlying processes such as circadian rhythm \citep{sancar2008intelligent}, cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, pattern formation, and metabolism  \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. Despite a symmetry in functionality or phenotype these systems often differ, sometimes substantially, at the molecular level. How many different mechanisms have the same function? We urge the reader to consider the metaphorical black box.

  Gene regulatory networks with identical phenotypes (external dynamics) do not necessarily have identical kryptotypes (internal dynamics). Any linear and minimal system (a gene network) -- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes -- has an identical phenotype up to a change of coordinates. 
  \begin{align}
    h(t) &= C e^{A t} B \\
    &= C V^{-1} V e^{A t} V^{-1} V B \\
    &= C V^{-1} e^{V A V^{-1} t} V B \\
    &= \bar{C} e^{\bar{A} t} \bar{B}
  \end{align}
  Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same phenotype if they are related by a change of coordinates.

  Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
  \begin{align}
    VB &= B \implies \bar{B} = B \\
    CV &= C \implies \bar{C} = C
  \end{align}
  In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.

  Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
  \begin{align}
    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
  \end{align}
  and a \emph{Linear Evolutionary System} is, 
  \begin{align}
    \mathcal{S}(\tau) := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
  \end{align}
  That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\mathcal{S}(\tau)$.

  More generally, \plr{introduce Kalman},
  we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
  \begin{equation} \label{eqn:equivalence}
      \begin{aligned}
      \calA_n(A_0) 
          &= \left\{
              A : C e^{At} B = C e^{A_0 t} B \; \text{for}\; t \ge 0 
          \right\}  \\
          &= \left\{
              A : C A^k B = C A_0^k B \; \text{for}\; 1 \le k \le n-1 
          \right\} .
      \end{aligned}
  \end{equation}
  Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
  Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

  Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$. This set can be completely parameterized using the \emph{Kalman Decomposition}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined. Note that this implies that there is always more than one possible gene regulatory network architecture per phenotype, however we can analytically describe the complete set.

  \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks]

    Let
    \begin{align*}
      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
    \end{align*}
    with $V(\tau)$ preserving both $B$ and $C$, then
    the set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in Example \ref{ex:oscillator} are given by
    \begin{align*}
      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
\end{scope}

\end{tikzpicture}
    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any value}
    \end{figure}
 % \end{example}

 % \begin{example}[External Equivalence does not imply internal equivalence]
   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the internal dynamics, or kryptotypes, vary as a function of $\tau$. 
    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
    \caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom). Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
  \end{figure}
  \end{example}

  %%%%%%%%%%%%%%%%%%%%%%%%
  \section*{Intraspecific variation and genetic drift}

%  \plr{rework this: we assume optimal phenotype is fixed and population is large but polymorphic and not strictly on the optimum}

%First we focus on a simple evolutionary scenario -- \emph{phenotypically invariant evolution}: a large population, perfectly adapted, and in a constant environment. In this circumstance we expect phenotype to be conserved throughout evolutionary time. As such we should only expect the phenotype to change as a consequence of genetic drift (small effective population size), adaptation to new selective and/or environemental pressures. These phenotypic variations should yield distinct signatures. Adaptative changes will change the optimal impulse response function $h(t) \xrightarrow{adaptation} h'(t)$. Genetic drift registers as an increase in the intrapopulation variation in $h(t)$. 
% \jss{added the phenotypic invariance stuff here (above) -- not sure if it should put elsewhere yet}

  At any given time, there will be a range of network coefficients present in the population
  due to segregating genetic polymorphism.
  Over many generations, even if selective pressures do not change,
  this range of networks will shift 
  as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
  How much variation do we expect to find within a population?
  Is this range limited by available variation or kept in check by selection?
  How fast will a population explore the space of equivalent networks?
  To answer these questions, we need to know two things known at best only roughly \citep{felsenstein1988phylogenies} --
  the strength of stabilizing selection on the phenotype,
  and the amount (and structure) of heritable variation in the genotype.
  We aim to get order-of-magnitude estimates.

  \plr{amount of heritable variation}
  We quantify (roughly) the amount of heritable variation
  by $\sigma^2$, the genetic variance present in a population in a typical entry of $A$.
  The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
  per change in concentration of $j$.
  It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
  than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
  In the context of transcription factor networks this may be affected 
  not only by the binding strength of $j$ to the promoter region of $i$
  but also the effects of other transcription factors (e.g., cooperativity)
  and local chromatin accessibility \citep{stefflova2013cooperativity}.
  For this reason, 
  the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
  typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region.
  (Recall that although these are best modeled through nonlinear terms,
  by linearizing we essentially consider first-order effects.)
  On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
  in the presence of substantial molecular noise \citep{canalization,buffering},
  suggesting that substantial variation in the micro-scale dynamics we consider here
  may be necessary to produce relevant phenotypic effects downstream.
  \plr{replace "micro-scale" with sthg else or discuss earlier}
  Differences in $A_{ij}$ due to a sequence change are hard to measure 
  -- the closest available data we could find either related to variation in transcription factor binding site occupancy
  or in expression levels (e.g., cis-eQTL).
  The former may be an overestimate, 
  since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin)
  and the latter probably measures changes in steady-state concentration (our $x_i$) rather than the rate of change.
  \citet{kasowski2010variation} found differential occupancy in 7.5\%
  of binding sites of a transcription factor (p65) between human individuals.
  \citet{verlaan2009targeted} showed that cis-regulatory variation
  accounts for around 2--6\% of expression variation in human blood-derived primary cells,
  while \citep{lappalainen2013transcriptome} found that human population variation 
  explained about 3\% of expression variation.
  \plr{Get some data from at least one other species in here!}
  Taken together, this suggests that variation in the entries of $A$
  may be on the scale of 1\% between individuals of a population --
  doubtless varying substantially between species and between genes.

  \plr{merge this above}
  The amount and structure of this standing variation is established over long time scales
  by many factors, including
  mutation-selection balance, 
  shifts in the phenotypic optimum,
  and/or spatial variation in the optimum \citep{hansen1996translating}.
  Quantitative genetics models of mutation-selection balance 
  predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
  but it is unclear how well these predictions match reality \citep{johnson_barton}
  and how much they are expected to change over time \citep{arnold_changing_G}.
  However, empirical work allows us to estimate at least the rough magnitude of variation.

  \plr{strength of selection}
  It seems certain that selection is not so strong that intra-population variation
  is strongly deleterious,
  so that if $u$ is the typical scale on which selection acts
  -- e.g., a phenotype that differs from the optimum by $x$ has fitness of $\exp(-x^2/2u^2)$ --
  then $u > \sigma$.
  However,
  a range of studies \plr{find them} have found evidence for weak stabilizing selection
  on regulatory SNPs and cis-eQTL.
  For instance, 
  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
  This suggests that 
  the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
  so that perhaps $\sigma$ and $u$ are relatively close.
  This is as would be expected if available variation is held in check by mutation--selection balance
  rather than genetic drift.
  A conservative estimate would be that $u = 5 \sigma$;
  taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
  to effect a noticeable drop in fitness.
  \plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}

>>>>

\section{Kalman Decomposition} \label{apx:kalman}
\begin{definition}[Phenotypic equivalence of systems]
    Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:lti_system}
    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
    and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
    \textbf{phenotypically equivalent} 
    if
    \begin{align*}
        \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
    \end{align*}
    Equivalently, this occurs if and only if
    \begin{align*}
        h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
    \end{align*}
    where $h$ and $\bar h$ are the impulse responses of the two systems.
\end{definition}
